---
title: "MDanilovaite_ch15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## 15.6 4 Exercise

Task:
The New York Times (January 8, 2003, page A12) reported the following
data on death sentencing and race, from a study in Maryland.

Analyze the data using the tools from this chapter. Interpret the results.
Explain why, based only on this information, you can’t make causal
conclusions. (The authors of the study did use much more information
in their full report.)

Solution:
1. With provided data, initial conclusion could be made that race influences death sentencing.
To test this, hypotheses to check for independence were formed:
  H0 : Race and Death sentence are independant
  H1 : Race and Death sentence are not independent
2. Likelihood ratio test statistic T was calculated, and Pearsons' test statistic for independence was performed

Calculated values:
T = 14.75
U = 14.96

p-value P(x^2 > T) = 0.0001
p-value P(x^2 > U) = 0.0001

H0 is rejected, as selected for of confidence interval = .95, p-value is 0.05, which is larger than calculated P(x^2 > T), P(x^2 > U)

Validity of rejection was confirmed with calculations of strength of dependence by estimating odds ratio and log-odds ratio. Then Wald statistic and p are:
W = 3.84
p-value P(|Z| > W) = 0.0001

H0 is rejected again.

3. as seen in step 2, H0 about variable independence was rejected, thus it could be infered, that race and death sentence are related. 
However, it is not clear under what conditions data was collected - e.g. was data collected from same prison or multiple (specific prisons could house non-violed offenders, while other could have over-represented violent offender population), or if crimes between sentenced population comparable. thus determined causal link based only on provided information cannot be used to make definite conclussions.

```{r 15_6_4, echo=FALSE}
df <- data.frame(matrix(ncol = 2, nrow = 0))
x <- c("Death Sentence", "No Death Sentence")
colnames(df) <- c("Death Sentence", "No Death Sentence")
df[nrow(df) + 1,] = c(90,165)#c(14,641)
df[nrow(df) + 1,] = c(84,307)#c(62,594)
row.names(df) <- c("Black Victim", "White Victim")

# ini variables
n <- 0
tmp <- 0
T <- 0
U <- 0
E <- 0
f_dregrees <- (nrow(df)-1)*(ncol(df)-1)

# find n of all observations
for (i in c(1,2))
{
  for (j in c(1,2))
  {
    n <- n + df[i,j]
  }
}

# intermediate calculation for T
for (i in c(1,2))
{
  for (j in c(1,2))
  {
   tmp <- tmp + df[i,j] *
     log( (df[i,j]*n)/
          (sum(df[i,])*sum(df[,j]))
        )
   #print(tmp)
  }
}

# final T
T <- 2*tmp

tmp <- 0

# intermediate calculation for U
for (i in c(1,2))
{
  for (j in c(1,2))
  {
    E <- sum(df[i,])*sum(df[,j])/n
    tmp <- tmp + ((df[i,j]-E)^2)/E
    #print(tmp)
  }
}

# final U
U <- tmp

# one tailed test, Pearson Chi Square
print(round(T,2))
print(round(U,2))
print(round(pchisq(T, df=f_dregrees,lower.tail=FALSE),4))
print(round(pchisq(U, df=f_dregrees,lower.tail=FALSE),4))

# because degrees of freedom = 1, confidence interval = .95, p-value is 0.05
# this means that H0 of independence is rejected

# Uncoment to check calculated values with Pearson Chi square test
#chi <- chisq.test(df)
#print(chi)

#knitr::kable(head(df[,]), "pipe")

# strength of dependence calculation

odds_r <- (df[1,1]*df[2,2])/(df[1,2]*df[2,1])
l_odds_r <- log(odds_r)
se_l_odds_r <- sqrt((1/df[1,1])+(1/df[1,2])+(1/df[2,1])+(1/df[2,2]))
se_odds_r <- odds_r*se_l_odds_r

W <- l_odds_r/se_l_odds_r

print(round(W,2))
print(round(pnorm(W, lower.tail = FALSE),4))

```

## 15.6 5 Exercise

Task:
Analyze the data on the variables Age and Financial Status from:
https://www.stat.cmu.edu/~larry/all-of-statistics/=data/montana.dat

Solution:
1. Hipotheses about independence were formed:
  H0: age and financial status are independent
  H1: age and financial status are not independent
  
2. Calculated statistics:
T = 22.06
U = 20.68
p-value log likelihood = 0.0002
p-value Pearson = 0.0004

H0 was rejected, as calculated statistics when confidence interval = .95, and p-value is 0.05 for selected interval, are smaller that 0.05.
Thus, age and financial status correlates

```{r 15_6_5, echo=FALSE}
# read dataset and select columns:
#readLines("https://www.stat.cmu.edu/~larry/all-of-statistics/=data/montana.dat", n=50)
df <- read.table( "https://www.stat.cmu.edu/~larry/all-of-statistics/=data/montana.dat", header=TRUE, skip=31 )
df = subset(df, select = -c(SEX,INC,POL,AREA,STAT) )

# create contingency table:
df = table(df)
df<-as.data.frame.matrix(df)
df<-df[-c(1),]
df<-df[,-c(1)]

colnames(df) <- c("worse", "same","better")
row.names(df) <- c("under 35", "35-54", "55 and over")

print(df)

# two discrete variables analysis:
# ini variables
n <- 0
tmp <- 0
T <- 0
U <- 0
E <- 0
f_dregrees <- (nrow(df)-1)*(ncol(df)-1)

# find n of all observations
for (i in c(1,2,3))
{
  for (j in c(1,2,3))
  {
    n <- n + df[i,j]
  }
}

# intermediate calculation for T
for (i in c(1,2,3))
{
  for (j in c(1,2,3))
  {
   tmp <- tmp + df[i,j] *
     log( (df[i,j]*n)/
          (sum(df[i,])*sum(df[,j]))
        )
   #print(tmp)
  }
}

# final T
T <- 2*tmp

tmp <- 0

# intermediate calculation for U
for (i in c(1,2,3))
{
  for (j in c(1,2,3))
  {
    E <- sum(df[i,])*sum(df[,j])/n
    tmp <- tmp + ((df[i,j]-E)^2)/E
    #print(tmp)
  }
}

# final U
U <- tmp

# one tailed test, Pearson Chi Square
print(round(T,2))
print(round(U,2))
print(round(pchisq(T, df=f_dregrees,lower.tail=FALSE),4))
print(round(pchisq(U, df=f_dregrees,lower.tail=FALSE),4))

# because degrees of freedom = 4, confidence interval = .95, p-value is 0.05
# this means that H0 of independence is rejected

# uncomment to check calculated values with Pearson Chi square test
#chi <- chisq.test(df)
#print(chi)

#knitr::kable(head(df[,]), "pipe")
```

## 15.6 6 Exercise

Task:
Estimate the correlation between temperature and latitude using the
data from
https://www.stat.cmu.edu/~larry/all-of-statistics/=data/temp.dat
Use the correlation coefficient. Provide estimates, tests, and confidence
intervals.

Solution:

Coeficient estimates were calculated. Confidence intervals were calculated and are provided below:
Pearson correlation = -0.8480352
95% confidence interval = [-0.9084082, -0.7530174]

Spearman correlation = -0.8154035
95% confidence interval = [-0.8964244 -0.6816441]


```{r 15_6_6, echo=FALSE}
# ini libraries
if(!require(rcompanion)){
    install.packages("rcompanion")
    library(rcompanion)
}

# read dataset and select columns:
#readLines("https://www.stat.cmu.edu/~larry/all-of-statistics/=data/temp.dat", n=50)
df <- read.table( "https://www.stat.cmu.edu/~larry/all-of-statistics/=data/temp.dat", header = TRUE, fill = TRUE, sep = '\t', skip=30)
df = subset(df, select = -c(City,Long) )

# reshape df to fit 2 x n shape (chapter 14.2)
df <- t(df)
# data points
n <- ncol(df)

# ini variables
alpha_2 <- (1 - 0.95)/2
z_alpha_2 <- 1.96 # from https://www2.southeastern.edu/Academics/Faculty/dgurney/Math241/StatTopics/ZAlpha.htm

# calculate pearson correlation
rho <- cor(df[1,], df[2,],  method = "pearson", use = "complete.obs")

#print(ifisherz(fisherz(rho)))

# confdence interval with delta method
theta <- 0.5 * (log(1 + rho) - log(1 - rho))
se_theta <- 1 / sqrt(n-3)

a <- theta - (z_alpha_2 / sqrt(n-3))
b <- theta + (z_alpha_2 / sqrt(n-3))

rho_min <- (exp(2*a) - 1) / (exp(2*a) + 1)
rho_max <- (exp(2*b) - 1) / (exp(2*b) + 1)

print(rho) # pearson correlation
print(rho_min) # conf. interval lower bound
print(rho_max) # conf. interval upper bound

# confidence interval with bootstrap
spearman_rho = cor(df[1,], df[2,],  method = "spearman", use = "complete.obs")


print(spearman_rho) # spearman correlation

n <- sum(complete.cases(df[1,], df[2,]))

# bootstrapped confidence interval
print(sort(tanh(atanh(spearman_rho) + c(-1,1)*sqrt((1+spearman_rho^2/2)/(n-3))*qnorm(p = alpha_2))))


```
## 15.6 7 Exercise

Task:
Test whether calcium intake and drop in blood pressure are associated.
Use the data in
https://www.stat.cmu.edu/~larry/all-of-statistics/=data/calcium.dat

Solution:
1. Hipotheses were formed
  H0: calcium intake and blood pressure drop are independent
  H1: calcium intake and blood pressure drop are not independent
  
2. To test hipotheses, dataset was split into two subsets, to produce two  based on 15.12 theorem -> one continuous one discrete variable
http://www.sthda.com/english/wiki/one-way-anova-test-in-r

associated


```{r 15_6_7, echo=FALSE}
# read dataset and select columns:
#readLines("https://www.stat.cmu.edu/~larry/all-of-statistics/=data/calcium.dat", n=50)
df <- read.table( "https://www.stat.cmu.edu/~larry/all-of-statistics/=data/calcium.dat", header = TRUE, fill = TRUE, sep = '\t', skip=32)
df = subset(df, select = -c(Begin,End) )

# 15.12: =================================================

f1 = subset(df, df$Treatment == "Placebo")
f1 <- f1[,2]
f2 = subset(df, df$Treatment == "Calcium")
f2 <- f2[,2]

D_tmp <- rep(0, nrow(df)) 

for (i in 1:nrow(df))
  {
    f1_hat <- sum(f1 <= df[i,2]) / length(f1)
    f2_hat <- sum(f2 <= df[i,2]) / length(f2)
    D_tmp[i] = abs(f1_hat - f2_hat)
  }

D = max(D_tmp)
  
n1 = length(f1)
n2 = length(f2)
test_statistic = sqrt(n1 * n2 / (n1 + n2)) * D

# 
H_inv <- function(t, q = test_statistic, xtol = 1e-8)
{
  if (t != 0) 
  { "t must be non-zero"
    t2 = t * t
    j_max = round(sqrt(- log(xtol / 2) / (2 * t2) ), digits = 0)
    jj = seq(1, j_max+1, by=1)

    result <- ((1 + 2 * sum((-1)**(jj) * exp(-2 * (jj**2) * t2))) - q)**2
  }
}

h_inv_test_statistic <- optim(1.0, H_inv, method = "BFGS" )$par

print(cat('D: ',  D))
print(cat('Test statistic: ',  test_statistic))
print(cat('H^{-1}(test_statistic): ', h_inv_test_statistic))

# compare
# two sample Kolmogorov-Smirnov test 
ks.test(f1, f2)

# One way analysis of variance (one way ANOVA)
res.aov <- aov(Decrease ~ Treatment, data = df)
summary(res.aov)

```

