---
title: "MDanilovaite_ch22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## 22.13 3 Exercise

Download the spam data from:
http://www-stat.stanford.edu/∼tibs/ElemStatLearn/index.html
The data file can also be found on the course web page. The data contain 57 covariates relating to email messages. Each email message was
classified as spam (Y=1) or not spam (Y=0). The outcome Y is the last
column in the file. The goal is to predict whether an email is spam or
not.
(a) Construct classification rules using (i) LDA, (ii) QDA, (iii) logistic
regression, and (iv) a classification tree. For each, report the observed
misclassification error rate and construct a 2-by-2 table of the form

h(x)=0 
h(x)=1
Y = 0 ?? ??
Y = 1 ?? ??
(b) Use 5-fold cross-validation to estimate the prediction accuracy of
LDA and logistic regression.
(c) Sometimes it helps to reduce the number of covariates. One strategy
is to compare Xi for the spam and email group. For each of the 57
covariates, test whether the mean of the covariate is the same or different
between the two groups. Keep the 10 covariates with the smallest pvalues. Try LDA and logistic regression using only these 10 variables


```{r 22_13_3, echo=FALSE}
# read dataset and select columns:
library(MASS)
library(caret)
library(rpart)
library(dplyr)

# read dataset and select columns:
#readLines("https://www.stat.cmu.edu/~larry/all-of-statistics/=data/spam.dat", n=50)
df <- read.table( "https://www.stat.cmu.edu/~larry/all-of-statistics/=data/spam.dat", header = FALSE, fill = TRUE, sep = ' ')

# separate to training and test
set.seed(123)
indexes <- createDataPartition(df[,58], p = .70, list = F)
train <- df[indexes, ]
test <- df[-indexes, ]

# LDA classifier
fit_lda <- lda(V58~., data = train)
pred_lda <- predict(fit_lda, test[,-58])
confusionMatrix(pred_lda$class, as.factor(test$V58))

# QDA
fit_qda <- qda(V58~., data = train)
pred_qda <- predict(fit_qda, test[,-58])
confusionMatrix(pred_qda$class, as.factor(test$V58))

# Logistic Regression
fit_lr <- glm(V58 ~., data = train, family = "binomial")
pred_lr <- predict(fit_lr, test, type = "response")
pred_lr_rd <- ifelse(pred_lr > 0.5, 1, 0)
confusionMatrix(as.factor(pred_lr_rd), as.factor(test$V58))

# Classification tree
fit_tree <- rpart(V58 ~., data = train, method = 'class')
pred_tree <- predict(fit_tree, test, type = 'class')
confusionMatrix(pred_tree, as.factor(test$V58))

# Crossvalidation
ctrl <- trainControl(method = "cv", number = 5)

# LDA
model_lda <- train(train[,-58], as.factor(train[,58]), method = "lda", trControl = ctrl)

# Logistic regression
model_lr <- train(train[,-58], as.factor(train[,58]), method = "glm", trControl = ctrl)

set_1 <- subset(df, V58 == 1)
set_0 <- subset(df, V58 == 0)

tmp=c()

# Feature design
for (i in 1:(ncol(df)-1))
  {
    t <-t.test(set_0[,i], set_1[,i], alternative = "two.sided", var.equal = FALSE)

    tmp <- rbind(tmp, c(i, t$p.value))
  } 

tmp <- tmp[order(tmp[,2],decreasing = FALSE),]
tmp <- tmp[c(1:10),]

df_10 <- df[,c(tmp[,1],58)]
# separate to training and test
set.seed(123)
indexes <- createDataPartition(df_10[,11], p = .70, list = F)
train <- df_10[indexes, ]
test <- df_10[-indexes, ]

# LDA classifier
fit_lda <- lda(V58~., data = train)
pred_lda <- predict(fit_lda, test[,-11])
confusionMatrix(pred_lda$class, as.factor(test$V58))

# Logistic Regression
fit_lr <- glm(V58 ~., data = train, family = "binomial")
pred_lr <- predict(fit_lr, test, type = "response")
pred_lr_rd <- ifelse(pred_lr > 0.5, 1, 0)
confusionMatrix(as.factor(pred_lr_rd), as.factor(test$V58))
```
## 22.13 6 Exercise

Use VC theory to get a confidence interval on the true error rate of the
LDA classifier for the iris data (from the book web site) (does not exists anymore).

Number of training points that can be classified exactly is VC dimension:
 -Measures relevant size of hypothesis space, as with decision trees with k leaves

```{r 22_13_6, echo=FALSE}
library(MASS)
library(caret)

# read dataset and select columns:
readLines("https://www.stat.cmu.edu/~larry/all-of-statistics/=data/spam.dat", n=50)
df <- read.table( "https://www.stat.cmu.edu/~larry/all-of-statistics/=data/spam.dat", header = FALSE, fill = TRUE, sep = ' ')

# LDA classifier
set.seed(123)
indexes <- createDataPartition(df[,58], p = .90, list = F)
train <- df[indexes, ]
test <- df[-indexes, ]
fit_lda <- lda(V58~., data = train)
coef(fit_lda)
pred_lda <- predict(fit_lda, test[,-58])

# missclassification error rate
false_positives = sum((pred_lda$class == 1) & (test[,58] == 0))
false_negatives = sum((pred_lda$class == 0) & (test[,58] == 1))

L_h <- (false_positives + false_negatives) / nrow(test)

# declare variables
alpha <- 0.05
n <- nrow(df)
# vc for linear classifier
vc_dim <- ncol(df) # features + 1 -> in this case 58th column is class

# theorem 23.21
epsilon <- (32 / n) * log( 8 *((n ^ vc_dim) + 1) / alpha)
epsilon <- sqrt(epsilon)
print(epsilon)
print(L_h)
print(L_h - epsilon)
print(L_h + epsilon) # error interval too large -> rate should be in [0,1)

# reduced vc to 10
epsilon <- (32 / n) * log( 8 *((n ^ 10) + 1) / alpha)
epsilon <- sqrt(epsilon)
print(epsilon)
print(L_h)
print(L_h - epsilon)
print(L_h + epsilon) # conf interval too large -> rate should be in [0,1)

```
